Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-09-17 manuscript/climate_sensitivity_SI.Rmd
Added small updates to text and table
Small updates to text and table captions

bark <- read.csv("tables_figures/publication/tableS2_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
colnames(bark) <- c("Species", "Equations", "$R^2$")
kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific regression equations for bark thickness (mm) as a function of diameter at breast height without bark (mm).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```
We used linear regression on log-transformed data to relate $r_{bark}$ to the diameter inside bark from 2008 data. These were then used to determine $r_{bark}$ in the $DBH$ reconstruction. No bark correction was applied for *Fagus grandifolia*, which has thin bark.

@@

\newpage
```{r table S9, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS9_Rt_multivariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S9. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(8, bold = T) %>%
  row_spec(15, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.

\newpage
```{r table S10, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS10_Rc_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S10. Summary of top models for each drought instance, where $Rc$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 10, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 21, 26, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(10, bold = T) %>%
  row_spec(21, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and appeared in all its top models **(1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models)**.

\newpage
```{r table S11, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS11_Rs_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S11. Summary of top models for each drought instance, where $Rs$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 6, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 11, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 17, 21, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(6, bold = T) %>%
  row_spec(11, bold = T) %>%
  row_spec(17, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and appeared in all its top models
**(1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172)**.

>>>>

bark <- read.csv("tables_figures/publication/tableS2_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
colnames(bark) <- c("Species", "Equations", "$R^2$")
kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific regression equations for bark thickness (mm) as a function of diameter at breast height without bark (mm).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```
We used linear regression on log-transformed data to relate $r_{bark}$ to the diameter inside bark from 2008 data. These were then used to determine $r_{bark}$ in the $DBH_{Y}$ reconstruction (DBH in year Y). No bark correction was applied for *Fagus grandifolia*, which has thin bark.

@@

\newpage
```{r table S9, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS9_Rt_multivariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S9. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(8, bold = T) %>%
  row_spec(15, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).

\newpage
```{r table S10, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS10_Rc_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S10. Summary of top models for each drought instance, where $Rc$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 10, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 21, 26, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(10, bold = T) %>%
  row_spec(21, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and appeared in all its top models (1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models).

\newpage
```{r table S11, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS11_Rs_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
colnames(multivar_table)[1] <- "drought"
kable(multivar_table, booktabs = TRUE, caption = "Table S11. Summary of top models for each drought instance, where $Rs$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  pack_rows("", 1, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 6, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 11, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 17, 21, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(6, bold = T) %>%
  row_spec(11, bold = T) %>%
  row_spec(17, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
$R^2$ refers to conditional $R^2$. 
Year was included in the model for all drought years and appeared in all its top models (1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172).

### 2020-09-17 manuscript/climate_sensitivity_draft.Rmd
Small updates to text and table captions

In the context of the multivariate model, $Rt$ did not vary across drought years. That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9). 
In contrast, both $Rc$ and $Rs$ varied across years, being highest in all of the top models (Tables S10-S11), in which both were highest in 1966, intermediate in 1977, and lowest in 1999.

@@

Drought resistance and resilience, but not recovery, were negatively correlated with $PLA_{dry}$ and $\pi_{tlp}$ (Fig. **4**; Tables 1, S4-S11). 
For $Rt$, $PLA_{dry}$ had a significant influence, with negative coefficient, in top models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). It was also included in some of the top models for 1999 (Tables S8-S9). 
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S8), although its influence was not significant at $\Delta$AICc<2. 
It was also included in some of the top models for 1999 (Tables S8-S9).

@@

For $Rs$, $PLA_{dry}$ was in the best models for the three droughts combined and for the 1966 drought individually, and some of the top models for 1977 and 1999 (Fig. **4**; Table S11); however, its effects were not significant atat $\Delta$AICc<2. 
$\pi_{tlp}$ was in the best models for the three droughts combined and for 1966 and 1999 individually, and in one of the top models for 1977 (Fig. **4**; Table S11). It's effects were significant at $\Delta$AICc<2 for 1999 only.

@@

Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*. 
Mechanistically, this is consistent with, and reinforces, previous findings that it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016]. 
Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

>>>>

In the context of the multivariate model, $Rt$ did not vary across drought years. That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9). 
In contrast, "drought year" varied for both $Rc$ and $Rs$, and were present in all of the top models for the combined scenario (Tables S10-S11). Between both metrics, coefficients for "drought year" were highest in 1966, intermediate in 1977, and lowest in 1999.

@@

Drought resistance and resilience, but not recovery, were negatively correlated with $PLA_{dry}$ and $\pi_{tlp}$ (Fig. **4**; Tables 1, S4-S11). 
For $Rt$, $PLA_{dry}$ had a significant influence, with negative coefficient, in top models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). It was also included in some of the top models for 1999 (Tables S8-S9). 
$\pi_{tlp}$ was included with a negative coefficient in the best model for the combined droughts scenario and for the 1977 drought individually (Fig. **4**; Table S8), although its influence was not significant at $\Delta$AICc<2. 
It was also included in some of the top models for 1999 (Tables S8-S9).

@@

For $Rs$, $PLA_{dry}$ was in the best models for the three droughts combined and for the 1966 drought individually, and some of the top models for 1977 and 1999 (Fig. **4**; Table S11); however, its effects were not significant at $\Delta$AICc<2. 
$\pi_{tlp}$ was in the best models for the three droughts combined and for 1966 and 1999 individually, and in one of the top models for 1977 (Fig. **4**; Table S11). However, its effects were significant at $\Delta$AICc<2 for 1999 only.

@@

Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*. 
Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016]. 
Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
